How fluid-mechanical erosion creates anisotropic porous media

### 2022-06-23 docs/PhysicsPaper/StokesErosion.tex
NM

For simplicity, we will assume $\bvec{k}  = \diag(k_{11}, k_{22})$ to be a diagonal matrix for the sake of characterizing the permeability of the porous medium. Because the diagonal components of $\bvec{k}$ need not be equal, the medium can have different permeabilities in the horizontal and vertical directions. Further, we will assume that $\bvec{k}$ to be spatially homogeneous in order to characterize the medium with a single bulk quantity at any instance in time. Naturally, the permeability will change over time as the bodies that comprise the medium erode. Consider first the horizontal component of \eqref{eqn:Darcy}
\begin{equation}
\label{q1}
q_1(x,y) = -k_{11} \pderiv{p}{x}(x,y)
\end{equation}

@@

Projecting \eqref{divthm} onto $\ex$, using \eqref{eqn:drag}, and rearranging gives the exact relationship
\begin{equation}
\label{drag_press}
\FD \cdot \ex = \int_{-1}^{1} \left( -p(x,y) + 2 \mu u_x(x,y) \right) \big|_{x=-x_0}^{x=x_0} dy + 
\int_{-x_0}^{x_0} \mu u_y(x,y) \big|_{y=-1}^{y=1} dx
\end{equation}
We will now make some simplifying assumptions. First, because the imposed flow profile is uniform $\UU = (U,0)$, slip is permitted along the top and bottom boundaries $y=\pm 1$ and so the viscous stress will generally be much smaller than on the no-slip erodable boundaries. We will therefore drop the contribution from the second integral above. Second, if $x_0 > 1$ is chosen a sufficient distance from the erodable bodies, the flow profile will approximately match the uniform profile, implying that the term involving $u_x(\pm x_0, y)$ can be dropped. In addition, the pressure $p(\mp x_0, y)$ will approximately match the upstream and downstream values respectively. With these assumptions, \eqref{drag_press} simplifies to the approximate form
\begin{equation}
\FD \cdot \ex \approx 2 (\pup - \pdn)
\end{equation}
Combining with \eqref{eqn:k11} yields a formula relating the drag and permeability in the longitudinal direction
\begin{equation}
\label{perm_drag}
\frac{1}{k_{11}} \approx \frac{1}{4 U}  \, \FD \cdot \ex
\end{equation}
This formula serves as a valuable benchmark linking the microscopic (Stokes) and macroscopic (Darcy) perspectives. Since the drag is a quantity that can be computed with high accuracy with our Stokes solvers, verifying the relationship \eqref{perm_drag}, as will be done in Fig.~\ref{fig2}, validates a number of approximations made in the coarse-grained approach, such as (1) the homogenization of the flow field implicit in Darcy's law \eqref{eqn:Darcy}; (2) the calculation of the permeability from measurements of the flow rate and upstream/downstream pressures through \eqref{eqn:k11}.

@@

To begin, we discuss a single simulation of 80 bodies eroding in Stokes flow. For reference, the simulation discussed here is the same one shown in Fig.~\ref{fig1}. The initial configuration consists of 80 circular bodies having randomized sizes and positions. The material-removal process alters both the shape of individual bodies as well as the overall structure of the pore network transmitting the flow, as can be seen in Fig.~\ref{fig1}. The initial porosity is $0.4$, which corresponds to a dense packing (see Fig.~\ref{fig1} for a visual), and Fig.~\ref{fig2}(a) shows how the porosity increases over time as the bodies vanish. 
The resistance to flow decreases with this increasing porosity, as shown in Fig.~\ref{fig2}(b)

@@

Text for Fig 3: 
The anisotropy of permeability grows large, up to a value of 4 during the middle portion of the simulation. The figure also shows the anisotropy of the circular grid.

>>>>

For simplicity, we assume $\bvec{k}$ to be a diagonal matrix, $\bvec{k}  = \diag(k_{11}, k_{22})$, for the sake of characterizing the permeability of the porous medium. Because the diagonal components of $\bvec{k}$ need not be equal, the medium can have different permeabilities in the longitudinal and transverse directions. Further, we will assume $\bvec{k}$ to be spatially homogeneous in order to characterize the medium with a single bulk quantity at any instance in time. Naturally, the permeability will change over time as the bodies that comprise the medium disintegrate. Consider first the horizontal component of \eqref{eqn:Darcy}
\begin{equation}
\label{q1}
q_1(x,y) = -k_{11} \pderiv{p}{x}(x,y)
\end{equation}

@@

Projecting \eqref{divthm} onto the horizontal direction $\ex$, using \eqref{eqn:drag}, and rearranging gives the exact relationship
\begin{equation}
\label{drag_press}
\FD \cdot \ex = \int_{-1}^{1} \left( -p(x,y) + 2 \mu u_x(x,y) \right) \big|_{x=-x_0}^{x=x_0} dy + 
\int_{-x_0}^{x_0} \mu u_y(x,y) \big|_{y=-1}^{y=1} dx
\end{equation}
We will now make some simplifying assumptions. First, because the imposed flow profile is uniform $\UU = (U,0)$, slip is permitted along the top and bottom boundaries $y=\pm 1$. Therefore the viscous stress is generally much smaller along these top and bottom surfaces than on the no-slip erodable boundaries. We therefore drop the contribution from the second integral above. Second, if $x_0 > 1$ is chosen a sufficient distance from the erodable bodies, the flow profile approximately matches the uniform profile, implying that the term involving $u_x(\pm x_0, y)$ can be dropped. In addition, the pressure $p(\pm x_0, y)$ approximately matches the downstream and upstream values respectively. With these assumptions, \eqref{drag_press} simplifies to the approximate form
\begin{equation}
\FD \cdot \ex \approx 2 (\pup - \pdn)
\end{equation}
Combining with \eqref{eqn:k11} yields a formula relating the longitudinal drag and permeability
\begin{equation}
\label{perm_drag}
\frac{1}{k_{11}} \approx \frac{1}{4 U}  \, \FD \cdot \ex
\end{equation}
This formula establishes an important link between the microscopic (Stokes) perspective and the macroscopic (Darcy) perspective, and it gives us a way to test the assumptions involved in coarse-graining the system to extract porous-medium properties. That is, the total drag is an unambiguous quantity that can be computed with high accuracy in our Stokes-based simulations. The extraction of permeability, on the other hand, requires a few key approximations and assumptions, for example that the length-scale of grains is sufficiently small compared to the domain scale so that the flow field can be homogenized. Verifying relationship \eqref{perm_drag}, as will be done in Section \ref{sec:results}, will therefore support the idea that the collection of bodies can be treated as a porous medium and meaningful bulk quantities can be extracted.

%verifying the relationship \eqref{perm_drag}, as will be done in Fig.~\ref{fig2}, validates a number of approximations made in the coarse-grained approach, such as (1) the homogenization of the flow field implicit in Darcy's law \eqref{eqn:Darcy}; (2) the calculation of the permeability from measurements of the flow rate and upstream/downstream pressures through \eqref{eqn:k11}.

@@

To begin, we discuss a single simulation of 80 bodies eroding in Stokes flow. For reference, the simulation discussed here is the same one shown in Fig.~\ref{fig1}. The initial configuration consists of 80 circular bodies having randomized sizes and positions. The material-removal process alters both the shape of individual bodies as well as the overall structure of the pore network transmitting the flow, as can be seen in Fig.~\ref{fig1}. Initially, the solid bodies occupy $60\%$ of the area, or the porosity is $\varphi = 0.4$, which corresponds to a dense packing; see Fig.~\ref{fig1} for a visual. Figure~\ref{fig2}(a) shows how the solid-body area decreases, or equivalently the porosity increases, over time as the bodies erode. The increasing porosity serves as a convenient proxy for dimensionless time that is insensitive to the end conditions that set the strength of the flow (i.e.~whether we specify the pressure drop or the end velocity to be constant in time). We will henceforth use porosity, $\varphi$, on the horizontal axis of many plots to represent increasing time.

As the bodies disintegrate and give way to wider pores, the resistance to flow decreases as seen in Fig.~\ref{fig2}(b). The resistance to flow can be quantified in the two independent ways discussed in Section \ref{sec:medium}, namely: (1) We can measure the pressure drop, $\pup - \pdn$, across the cell and the total flux, $q_1$, to infer the {\em resistivity} or inverse permeability $1/k_{11}$; or (2) We can measure the combined drag on the collection of bodies. Figure \ref{fig2}(b) shows that these two approaches yield consistent results...

@@

%Text for Fig 3: 
%The anisotropy of permeability grows large, up to a value of 4 during the middle portion of the simulation. The figure also shows the anisotropy of the circular grid.
